Semantic Linearity in a Lazy Optimising Compiler

### 2024-01-26 icfp.tex
passed up to section 3.6.2

Finally, we consider the default case alternatives, also known as wildcards
(written $\_$), in the presence of linearity: matching against the wildcard
doesn't provide new information, so linearity is seen as before but without
fully consuming the scrutinee linear resources (in non-linear patterns) nor
binding new linear resources (in linear patterns). In short, if the scrutinee
is in WHNF, we can either use the resources from the scrutinee or the case
binder in that alternative. If the scrutinee is not in WHNF, we \emph{must} use
the case binder, as it's the only way to linearly consume the result of
evaluating the scrutinee. We also discuss the special case of a case
expression scrutinizing a variable $x$:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though the program is linear:
%begin{itemize}
% \item
if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
returning $x$ returns the result of evaluating $x$ that still had to be
consumed exactly once.
%\end{itemize}
However, in practice, it depends on the evaluation strategy. If linear function
applications are $\beta$-reduced \emph{call-by-name} (a common practice, as
linear functions use their argument exactly once), and the above function is
considered linear, then an application might duplicate linear resources during
evaluation. For example:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
Therefore, the type system we present in the next section, which evaluates
linear applications call-by-name, does not accept the above program as linear.
This interaction between evaluation and linearity
comes up in the type preservation proof of our program, and is again explored
with the reverse binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.

@@

In this chapter, we develop a linear calculus $\lambda_\Delta^\pi$, dubbed \emph{Linear Core}, that 
combines the linearity-in-the-arrow and multiplicity polymorphism introduced by
Linear Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
language, except for type equality coercions\footnote{We explain a main avenue of
future work, multiplicity coercions, in Section~\ref{sec:future-work}}.
%
Specifically, our core calculus is a linear lambda calculus with algebraic
datatypes, case expressions, recursive let bindings, and multiplicity
polymorphism.

Linear Core makes much more precise the various insights discussed in the
previous chapter by crystallizing them together in a linear type system for which we
prove soundness via the usual preservation and progress theorems. Crucially,
the Linear Core type system accepts all the \emph{semantically linear} example
programs (highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
Besides type safety, we prove that multiple optimising Core-to-Core
transformations preserve linearity in Linear Core. These same transformations
don't preserve linearity under Core's current type system. As far as we know,
we are the first to prove optimisations preserve types in a non-strict linear
language.

The first key idea for typing linearity semantically is to delay \emph{consuming a
resource} to when a computation that depends on that resource is effectively
evaluated or returned, by annotating relevant binders with \emph{usage
environments} (\S~\ref{sec:usage-environments}).
%
The second key idea is to have two distinct rules for
case expressions, branching on whether the scrutinee is in Weak Head Normal
Form, and using ``proof irrelevance'' to track resources that are no
longer available but haven't yet been fully consumed (\S~\ref{sec:lc-case-exps}). Additionally, we introduce
tagged resources to split the resources between the pattern-bound variables as
usage environments, encoding that pattern variables jointly ``finish
consuming'' the scrutinee resources.
% 
We also note that despite the focus on GHC Core, the fundamental ideas for
understanding linearity in a call-by-need calculus can be readily applied to
other non-strict languages.

We present Linear Core's syntax and type system iteratively, starting with the
judgements and base linear calculi rules for multiplicity and term lambdas plus
the variable rules ($\S~\ref{sec:base-calculi}$).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings ($\S~\ref{sec:usage-environments}$).
%
Finally, we introduce the rules to typecheck case expressions and alternatives,
along with the key insights to do so, namely branching on WHNF-ness of
scrutinee, proof irrelevant resources, and tagged variables
($\S~\ref{sec:lc-case-exps}$).

@@

The complete syntax of Linear Core is given by Figure~\ref{fig:full-linear-core-syntax}.
% Figure~\ref{fig:linear-core-types} and Figure~\ref{fig:linear-core-terms}.
%
The types of Linear Core are algebraic datatypes, function types, and
multiplicity schemes to support multiplicity polymorphism: datatypes
($T~\ov{p}$) are parametrised by multiplicities, function types
($\vp\to_\pi\s$) are also annotated with a multiplicity, which can be $1$,
$\omega$ (read \emph{many}), or a multiplicity variable $p$ introduced by a
multiplicity universal scheme ($\forall p.~\vp$).
%
\[
\SyntaxTypes
\]
%
The terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is guaranteed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.
\[
\SyntaxTerms
\]
Linear Core takes the idea of annotating lets with usage environments from the
unpublished Linear Mini-Core document by Spiwack et.
al~\cite{cite:minicore}, which first tentatively tackled Core's
linearity issues. We discuss this work in more detail in
Section~\ref{sec:linear-mini-core}.

@@

The (small-step) operational semantics of Linear Core are given by
Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
evaluation for Linear Core as it captures the non-strict semantics in which
our type system understands linearity, while being simpler to reason about than
call-by-need operational semantics.
% which is traditionally modelled with a
% mutable heap to store \emph{thunks} and the values they are overwritten with.
Furthermore, linear function applications, even in a \emph{call-by-need} system, are
usually reduced \emph{call-by-name} as the function argument is guaranteed to
be used exactly once (thus avoiding unnecessarily allocating memory on the heap
for a redundant \emph{thunk}).
%
Specifically, function applications are reduced by the standard
\emph{call-by-name} $\beta$-reduction, substituting the argument whole by
occurrences of the lambda binder in its body, case expressions evaluate their
scrutinee to WHNF, substituting the result by the case binder and constructor
arguments for pattern-bound bound variables matching on that same constructor.

@@

\subsection{Typing Foundations\label{sec:base-calculi}}

Linear Core ($\lambda^\pi_\Delta$) is a linear lambda calculus akin to Linear
Haskell's $\lambda^q_\to$ in that both have multiplicity polymorphism,
(recursive) let bindings, case expressions, and algebraic data types.
$\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,
case expressions, and alternatives (in its purpose to type semantic linearity).
%, and secondarily in only treating multiplicity polymorphism superficially.
%
% \todo{And our treatment of multiplicity polymorphism completely ignores
% algebraic treatment of multiplicities with semiring operations!!! Would it be
% sufficient to add a rule for application of variable multiplicity functions?}
%
Otherwise, the base rules of the calculus for, multiplicity and term,
abstraction and application are quite similar. In this section, we present
the linear calculi's typing rules that share much in common with
$\lambda^q_\to$, and, in the subsequent sections, the rules encoding the
novel insights from Linear Core in typing semantic linearity, that were
explored by example in Section~\ref{sec:linearity-semantically}.
%
We note, however, that we handle multiplicity polymorphism differently from
Linear Haskell in ignoring the multiplicity semiring and, instead,
conservatively treating all multiplicity polymorphic functions as linear.
%
The full type system is given by Figure~\ref{fig:linear-core-typing-rules},
with auxiliary judgements given by
Figure~\ref{fig:linear-core-other-judgements}.

@@

We start with the main typing judgement. As is customary for linear type
systems, we use two typing environments, an \emph{unrestricted} $\G$ and
a \emph{linear} $\D$ environment.
%
Variables in $\G$ can be freely discarded (\emph{weakened}) and duplicated
(\emph{contracted}), while resources in $\D$ must be used exactly once (hence
can't be discarded nor duplicated). Despite not having explicit weakening and
contraction rules in our system, they are available as admissible rules for
$\G$ (but not for $\D$), since, equivalently (via
\cite{91621fae-5e53-3497-8291-32b2fab5a743}), resources from $\G$ are
duplicated for sub-derivations and may unrestrictedly exist in the variable
rules.
% TODO: Likely rewrite that
%
The main typing judgement, written $\G;\D \vdash e : \s \to \vp$,
denotes that expression $e$ has type $\s \to \vp$ under
the unrestricted environment $\G$ and linear environment $\D$.

Occurrences of unrestricted variables from $\G$ are well-typed as long as the linear
environment is empty, while occurrences of linear variables are only well-typed
when the variable being typed is the only resource available in the linear context.
\[
\begin{array}{cc}
\TypeVarOmega & \TypeLinearVar
\end{array}
\]
In both cases, the linear context must contain exactly what is required to
prove the proposition, whereas the unrestricted context may contain arbitrary
variables.
%
Variables in contexts are annotated with their type and multiplicity, so
$\x[\pi]$ is a variable named $x$ of type $\s$ and multiplicity $\pi$.

Linear functions are introduced via the function type ($\s \to_\pi \vp$) with
$\pi = 1$, i.e. a function of type $\s \to_1 \vp$ (or, equivalently, $\s \lolli \vp$)
introduces a linear resource of type $\s$ in the linear environment $\D$ to then type an expression of type $\varphi$.
%
Unrestricted functions are introduced via the function type ($\s \to_\pi \vp$) with $\pi =
\omega$, and the $\lambda$-bound variable is introduced in $\G$:
\[
\begin{array}{cc}
\TypeLamIntroL & \TypeLamIntroW
\end{array}
\]
A linear function application is well-typed if there exists a disjoint split of the
linear resources into $\D,\D'$ s.t. the function and argument, each under a
distinct split, are both well-typed and the argument type matches the
function's expected argument type. Conversely, unrestricted resources are
duplicated and available whole to both sub-derivations.
\[
\TypeLamElimL \quad \TypeLamElimW
\]
An unrestricted function, unlike a linear one, consumes its argument
unrestrictedly (zero or more times). Therefore, in an unrestricted function
application, allowing any linear resources to occur in the argument expression
entails consuming those resources unrestrictedly, since the
variable binding the argument expression could be discarded or used more than
once in the function body. Thus, argument expressions to unrestricted functions must also
be unrestricted, i.e. no linear variables can be used to type them.
%
Typing linear and unrestricted function applications separately is less general
than typing applications of functions of any multiplicity $\pi$ by scaling (per the multiplicity
semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
linearity does not benefit much from doing so, and by keeping the simple approach we can have
the linear and unrestricted environments be separate.
% \todo{We might only need multiplicities for the case-of-case definition as it
% exists in ghc?, even then couldn't we do semiring scaling without variables?
% Also, dislike sentence}

Multiplicity abstractions ($\Lambda p.~e$) introduce a multiplicity variable
$p$ (in the unrestricted context), and construct expressions of type
$\forall p.~\dots$, i.e. a type universally quantified over a multiplicity
variable $p$. We note that, in the body of the abstraction, function types annotated
with a $p$ variable and datatype fields with multiplicity $p$ are typed as
thought they are linear functions and linear fields, because $p$ can be
instantiated at both $\omega$ and $1$.
\[
\TypeMultLamIntro \quad \TypeMultLamElim
\]
A multiplicity application instantiates a multiplicity-polymorphic type
$\forall p.~\s$ at a particular (argument) multiplicity $\pi$, resulting in an
expression of type $\s$ where occurrences of $p$ are substituted by $\pi$, i.e.
$\s[\pi/p]$.
The rule additionally requires that $\pi$ be \emph{well-formed} in order
for the expression to be well-typed, using the judgement $\G \vdash_{mult}
\pi$, where well-formedness is given by $\pi$ either being $1$, $\omega$, or an
in-scope  multiplicity variable in $\G$.
\[
\TypeWellFormedMult
\]

These rules conclude the foundations of our linear calculi. In subsequent
sections we type (recursive) let bindings and case expressions,
accounting for semantic linearity as per the insights construed in
Section~\ref{sec:linearity-semantically}, effectively distilling them into the
key ideas of our work -- encoded as rules.

@@

A \emph{usage environment} $\Delta$ is the means to encode the idea that lazy
bindings don't consume the resources required by the bound expression when
defined, but rather when the bindings themselves are fully consumed.
Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
environment} to denote that consuming these variables equates to consuming the
resources in the usage environment $\D$ they are annotated with, where a usage
environment is essentially a multiset of linear resources. $\Delta$-bound
variables are introduced by a handful of constructs, namely, (recursive) let
binders, case binders, and case pattern variables. In the following example, as
per the insights into semantic linearity developed in
Section~\ref{sec:semantic-linearity-examples}, the resources required to
typecheck the body of the binder $u$, $x$ and $y$, are only used if the let-var
$u$ is consumed in the let-body $e$.  Accordingly, the usage environment of
the let-bound $u$ is $\{x,y\}$:
\[
f = \lambda \xl.~\lambda \y[1].~\llet{u = (x,y)}{e}
\]
Furthermore, usage environments guarantee that using a $\Delta$-bound variable
is mutually exclusive with directly using the resources it is annotated with --
using the $\Delta$-bound variable consumes all linear resources listed in its
usage environment, meaning they are no longer available for direct usage.
Dually, using the linear resources directly means they are no longer available
to consume through the usage environment of the $\Delta$-bound variable.

Finally, we note that usage environments bear a strong resemblance to the linear
typing environments to the right of the semicolon in the main typing judgement,
i.e. the environment with the linear resources required to type an expression.
%
In fact, usage environments and linear typing contexts differ only in that the
former are used to annotate variables, while the latter used to type
expressions. Yet, this distinction is slightly blurred after introducing how
typing environments can be moved to usage environments, or otherwise occurs in
rules relating the two.

@@

A $\Delta$-bound variable $u$ is a variable annotated with a usage environment $\Delta$. Crucially, for any $\Delta$-bound variable $u$:
%
%\begin{enumerate}
%\item
(1) using $u$ is equivalent to using all the linear resources in $\Delta$;
%\item $u$ can only be used once, since the $\D$ resources will not be available afterwards
%\item
(2) using $u$ is mutually exclusive with using the $\Delta$ resources it depends on elsehow;
% \item
(3) $u$ can be safely discarded as long as the resources in $\Delta$ are consumed elsehow.
%\end{enumerate}
%
Fortunately, since linear resources must be linearly split across
sub-derivations, (2) follows from (1) since consuming the linear resources
in $\Delta$ to type $u$ makes them unavailable in the context of any other
derivation. Therefore, expressions using these resources a second time, directly, or indirectly through the same
(or other) usage environment, is ill-typed, as the resources are
already allocated to the derivation of $u$. Similarly, (3) also follows from
(1), because if the linear resources aren't consumed in the $\Delta$-var
derivation, they must be consumed in an alternative derivation (or otherwise
the expression is ill-typed).
%
These observations all boil down to one typing rule for $\Delta$-bound
variables, which fundamentally encodes (1), implying the other two points:
\[
\TypeVarDelta
\]
The rule reads that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is exactly the resources in the usage environment of
that variable.
% ROMES:TODO: Dizer alguma coisa tipo (blurring the distinction between usage envs and typing envs)?

$\D$-variables are always introduced in $\Gamma$ since they can be discarded
and duplicated, despite multiple occurrences of the same $\Delta$-variable not
possibly being well-typed as, ultimately, it would imply non-linear usage of
linear resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

\subsubsection{Lazy let bindings}

In Section~\ref{sec:semantic-linearity-examples}, we discussed how linear
resources used in let-bound expressions are only consumed when the same let-bound
expressions are fully evaluated, i.e. linear resources required by let-bound
expressions are consumed lazily.
%
Moreover, resources from a let-bound expression cannot be used \emph{together}
with the variable binding, since said resources would end up being consumed
more than once, violating (semantic) linearity -- the binder has to be used in
mutual exclusion with the linear resources required to type the expression it
binds, and either \emph{must} be used, or we'd be discarding resources.

>>>>

Finally, we consider default case alternatives, also known as wildcards
(written $\_$), in the presence of linearity: matching against the wildcard
doesn't provide new information, so linearity is seen as before but without
fully consuming the scrutinee linear resources (in non-linear patterns) nor
binding new linear resources (in linear patterns). In short, if the scrutinee
is in WHNF, we can either use the resources from the scrutinee or the case
binder in that alternative. If the scrutinee is not in WHNF, we \emph{must} use
the case binder, as it's the only way to linearly consume the result of
evaluating the scrutinee. We also discuss the special case of a case
expression scrutinizing a variable $x$:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though the program is linear:
%begin{itemize}
% \item
if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
returning $x$ returns the result of evaluating $x$ that still had to be
consumed exactly once.
%\end{itemize}
However, in practice, it depends on the evaluation strategy. If linear function
applications are $\beta$-reduced \emph{call-by-name} (a common practice, as
linear functions use their argument exactly once), and the above function is
considered linear, then an application might duplicate linear resources during
evaluation. For example:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
Therefore, the type system we present in the next section, which evaluates
linear applications call-by-name, does not accept the above program as linear.
This interaction between evaluation and linearity
comes up in the type preservation proof of our program, and is again explored
with the reverse binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.

@@

In this section, we develop a linear calculus $\lambda_\Delta^\pi$,
dubbed \emph{Linear Core}, that combines the linearity-in-the-arrow
and multiplicity polymorphism introduced by Linear
Haskell~\cite{cite:linearhaskell}, featuring algebraic datatypes, case
expressions, recursive let bindings, and multiplicity polymorphism.
(i.e.,~all of Core's features with the exception of equality
coercions).  We discuss multiplicity coercions as future work in
Section~\ref{sec:future-work}.
%

Linear Core makes precise the various insights discussed in the
previous chapter by crystallizing them together in a linear type system for which we
prove soundness via the standard type preservation and progress theorems. Crucially,
Linear Core typing accepts all the \emph{semantically linear} examples
(highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
% Besides type safety, we prove that multiple optimising Core-to-Core
% transformations preserve linearity in Linear Core. These same transformations
% don't preserve linearity under Core's current type system. As far as we know,
% we are the first to prove optimisations preserve types in a non-strict linear
% language.

The first key idea for typing linearity semantically is to delay \emph{consuming a
resource} to when a computation that depends on that resource is effectively
evaluated or returned, by annotating relevant binders with \emph{usage
environments} (Section~\ref{sec:usage-environments}).
%
The second key idea is to have two distinct rules for case
expressions, branching on whether the scrutinee is in Weak Head Normal
Form, and using ``proof irrelevance'' to track resources that are no
longer available but haven't yet been fully consumed
(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tagged
resources to split the resources between the pattern-bound variables
as usage environments, encoding that pattern variables jointly
``finish consuming'' the scrutinee resources.
% 
We also note that despite the focus on GHC Core, the fundamental ideas for
understanding linearity in a call-by-need calculus can be readily applied to
other non-strict languages.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules ($\S~\ref{sec:base-calculi}$).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings ($\S~\ref{sec:usage-environments}$).
%
Finally, we introduce the rules to typecheck case expressions and alternatives,
along with the key insights to do so, namely branching on WHNF-ness of
scrutinee, proof irrelevant resources, and tagged variables
($\S~\ref{sec:lc-case-exps}$).

@@

The complete syntax of Linear Core is given in Figure~\ref{fig:full-linear-core-syntax}.
% Figure~\ref{fig:linear-core-types} and Figure~\ref{fig:linear-core-terms}.
%
The types of Linear Core are algebraic datatypes, function types, and
multiplicity schemes to support multiplicity polymorphism: datatypes
($T~\ov{p}$) are parametrised by multiplicities, function types
($\vp\to_\pi\s$) are also annotated with a multiplicity, which can be $1$,
$\omega$ (read \emph{many}), or a multiplicity variable $p$ introduced by a
multiplicity universal scheme ($\forall p.~\vp$).
%

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is guaranteed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

% Linear Core takes the idea of annotating lets with usage environments from the
% unpublished Linear Mini-Core document by Spiwack et.
% al~\cite{cite:minicore}, which first tentatively tackled Core's
% linearity issues. We discuss this work in more detail in
% Section~\ref{sec:linear-mini-core}.

@@

The (small-step) operational semantics of Linear Core are given in
Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
evaluation for Linear Core as it captures the non-strict semantics in which
our type system understands linearity, while being simpler to reason about than
call-by-need operational semantics.
% which is traditionally modelled with a
% mutable heap to store \emph{thunks} and the values they are overwritten with.
Furthermore, linear function applications, even in a \emph{call-by-need} system, are
usually reduced \emph{call-by-name} as the function argument is guaranteed to
be used exactly once (thus avoiding unnecessarily allocating memory on the heap
for a redundant \emph{thunk}).
%
Specifically, function applications are reduced by the standard
\emph{call-by-name} $\beta$-reduction, substituting the argument by
occurrences of the binder in its body, case expressions evaluate their
scrutinee to WHNF, substituting the result by the case binder and constructor
arguments for pattern-bound bound variables matching on that same constructor.

@@

\subsection{Base Typing System\label{sec:base-calculi}}

Linear Core ($\lambda^\pi_\Delta$) is a linear lambda calculus akin to Linear
Haskell's $\lambda^q_\to$ in that both have multiplicity polymorphism,
(recursive) let bindings, case expressions, and algebraic data types.
$\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,
case expressions, and alternatives (in its purpose to type semantic linearity).
%, and secondarily in only treating multiplicity polymorphism superficially.
%
% \todo{And our treatment of multiplicity polymorphism completely ignores
% algebraic treatment of multiplicities with semiring operations!!! Would it be
% sufficient to add a rule for application of variable multiplicity functions?}
%

The core rules of the calculus for, multiplicity and term,
abstraction and application are similar to that of $\lambda^q_\to$ and
mostly standard. 
We note that we handle multiplicity polymorphism differently from
Linear Haskell in ignoring the multiplicity semiring and, instead,
conservatively treating all multiplicity polymorphic functions as linear.
%
The full type system is given in Figure~\ref{fig:linear-core-typing-rules},
with auxiliary judgements given in
Figure~\ref{fig:linear-core-other-judgements}.

@@

We start with the main typing judgement.
As is common for linear type
systems, we use two typing environments, an \emph{unrestricted} $\G$ and
a \emph{linear} $\D$ environment.
Thus, the main judgement is written $\G;\D \vdash e :
\s$,
denoting that expression $e$ has type $\s$ under
the unrestricted environment $\G$ and linear environment $\D$.

%
Variables in $\G$ can be freely discarded (\emph{weakened}) and
duplicated (\emph{contracted}), while resources in $\D$ must be used
exactly once (hence can't be discarded nor duplicated). Despite not
having explicit weakening and contraction rules in our system, they
are available as admissible rules for $\G$ (but not for $\D$), since,
equivalently \cite{91621fae-5e53-3497-8291-32b2fab5a743}, resources
from $\G$ are duplicated for sub-derivations and may unrestrictedly
exist in the variable rules.
% TODO: Likely rewrite that
%

Occurrences of unrestricted variables from $\G$ are well-typed as long as the linear
environment is empty, while occurrences of linear variables are only well-typed
when the variable being typed is the only resource available in the
linear context (rules $Var_\omega$ and $Var_1$, respectively).
% \[
% \begin{array}{cc}
% \TypeVarOmega & \TypeLinearVar
% \end{array}
% \]
In both cases, the linear context must contain exactly what is required to
prove the proposition, whereas the unrestricted context may contain arbitrary
variables.
%
Variables in contexts are annotated with their type and multiplicity, so
$\x[\pi]$ is a variable named $x$ of type $\s$ and multiplicity $\pi$.

Linear functions are introduced via the function type ($\s \to_\pi \vp$) with
$\pi = 1$, i.e. a function of type $\s \to_1 \vp$ (or, equivalently, $\s \lolli \vp$)
introduces a linear resource of type $\s$ in the linear environment $\D$ to then type an expression of type $\varphi$.
%
Unrestricted functions are introduced via the function type ($\s \to_\pi \vp$) with $\pi =
\omega$, and the $\lambda$-bound variable is introduced in $\G$ (rules
$\lambda I_1$ and $\Lambda I$). Function application is standard, with
linear function application splitting the linear context in two
disjoint parts, one used to type the argument and the other for the
function.
Arguments of unrestricted functions must also be unrestricted, i.e. no
linear variables can be used to type them. 
% \[
% \begin{array}{cc}
% \TypeLamIntroL & \TypeLamIntroW
% \end{array}
% \]
% A linear function application is well-typed if there exists a disjoint split of the
% linear resources into $\D,\D'$ s.t. the function and argument, each under a
% distinct split, are both well-typed and the argument type matches the
% function's expected argument type. Conversely, unrestricted resources are
% duplicated and available whole to both sub-derivations.
% \[
% \TypeLamElimL \quad \TypeLamElimW
% \]
% An unrestricted function, unlike a linear one, consumes its argument
% unrestrictedly (zero or more times). Therefore, in an unrestricted function
% application, allowing any linear resources to occur in the argument expression
% entails consuming those resources unrestrictedly, since the
% variable binding the argument expression could be discarded or used more than
% once in the function body. Thus, argument expressions to unrestricted functions must also
% be unrestricted, i.e. no linear variables can be used to type them.
% %

Typing linear and unrestricted function applications separately is less general
than typing applications of functions of any multiplicity $\pi$ by scaling (per the multiplicity
semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
linearity does not benefit much from doing so and so we opt for a
simpler design. 
% \todo{We might only need multiplicities for the case-of-case definition as it
% exists in ghc?, even then couldn't we do semiring scaling without variables?
% Also, dislike sentence}

Multiplicity abstractions ($\Lambda p.~e$) introduce a multiplicity variable
$p$ (in the unrestricted context), and construct expressions of type
$\forall p.~\s$, i.e. a type universally quantified over a multiplicity
variable $p$. We note that, in the body of the abstraction, function types annotated
with a $p$ variable and datatype fields with multiplicity $p$ are typed as
thought they are linear functions and linear fields, because $p$ can be
instantiated at both $\omega$ and $1$.
\[
\TypeMultLamIntro \quad \TypeMultLamElim
\]
A multiplicity application instantiates a multiplicity-polymorphic type
$\forall p.~\s$ at a particular (argument) multiplicity $\pi$, resulting in an
expression of type $\s$ where occurrences of $p$ are substituted by $\pi$, i.e.
$\s[\pi/p]$.
The rule additionally requires that $\pi$ be \emph{well-formed} in order
for the expression to be well-typed, using the judgement $\G \vdash_{mult}
\pi$, where well-formedness is given by $\pi$ either being $1$, $\omega$, or an
in-scope  multiplicity variable in $\G$.
% \[
% \TypeWellFormedMult
% \]

% These rules conclude the foundations of our linear calculi. In subsequent
% sections we type (recursive) let bindings and case expressions,
% accounting for semantic linearity as per the insights construed in
% Section~\ref{sec:linearity-semantically}, effectively distilling them into the
% key ideas of our work -- encoded as rules.

@@

A \emph{usage environment} $\Delta$ is the means to encode the idea that lazy
bindings do not consume the resources required by the bound expression when
defined, but rather when the bindings themselves are fully consumed.
Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
environment} to denote that consuming these variables equates to consuming the
resources in the annotated usage environment $\D$ (i.e.,~essentially a multiset of linear resources).
%
$\Delta$-bound
variables are introduced by several of constructs: (recursive) let
binders, case binders, and case pattern variables. In the following example, as
per the insights developed in
Section~\ref{sec:semantic-linearity-examples}, the resources required to
typecheck the body of the binder $u$, $x$ and $y$, are only used if
the let-bound variable
$u$ is consumed in the let-body $e$.  Accordingly, the usage environment of
the let-bound $u$ is $\{x,y\}$:
\[
f = \lambda \xl.~\lambda \y[1].~\llet{u = (x,y)}{e}
\]
Furthermore, usage environments guarantee that using a $\Delta$-bound variable
is mutually exclusive with directly using the resources it is annotated with --
using the $\Delta$-bound variable consumes all linear resources listed in its
usage environment, meaning they are no longer available for direct usage.
Dually, using the linear resources directly means they are no longer available
to consume through the usage environment of the $\Delta$-bound variable.

Finally, we note that usage environments bear a strong resemblance to the linear
typing environments in the main typing judgement,
i.e. the environment with the linear resources required to type an expression.
%
In fact, usage environments and linear typing contexts differ only in that the
former are used to annotate variables, while the latter used to type
expressions.
% Yet, this distinction is slightly blurred after introducing how
% typing environments can be moved to usage environments, or otherwise occurs in
% rules relating the two.

@@

A $\Delta$-bound variable $u$ is a variable annotated with a usage environment $\Delta$. Crucially, for any $\Delta$-bound variable $u$:
%
%\begin{enumerate}
%\item
(1) using $u$ is equivalent to using all the linear resources in $\Delta$;
%\item $u$ can only be used once, since the $\D$ resources will not be available afterwards
%\item
(2) using $u$ is mutually exclusive with using the $\Delta$ resources it depends on;
% \item
(3) $u$ can be safely discarded as long as the resources in $\Delta$ are consumed.
%\end{enumerate}
%
Fortunately, since linear resources must be linearly split across
sub-derivations, (2) follows from (1) since consuming the linear
resources in $\Delta$ to type $u$ makes them unavailable in the
context of any other derivation. Therefore, expressions using these
resources a second time, directly, or indirectly through the same (or
other) usage environment, are ill-typed, as the resources are already
allocated to the derivation of $u$. Similarly, (3) also follows from
(1), because if the linear resources aren't consumed in the
$\Delta$-var derivation, they must be consumed in an alternative
derivation (or otherwise the expression is ill-typed).
%
These observations all lead to the typing rule for $\Delta$-bound
variables:

\vspace{-0.5cm}
\[
\TypeVarDelta
\]
The rule reads that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is exactly the resources in the usage environment of
that variable.
%
$\D$-variables are unrestricted 
since they can be discarded and duplicated, despite multiple
occurrences of the same $\Delta$-variable not being
well-typed as, ultimately, it would imply non-linear usage of linear
resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

\subsection{Lazy let bindings}

In Section~\ref{sec:semantic-linearity-examples}, we discussed how linear
resources used in let-bound expressions are only consumed when the same let-bound
expressions are fully evaluated, i.e. linear resources required by let-bound
expressions are consumed lazily.
%
Moreover, resources from a let-bound expression cannot be used \emph{together}
with the variable binding, since said resources would end up being consumed
more than once, violating (semantic) linearity -- the binder has to be used in
mutual exclusion with the linear resources required to type the expression it
binds, and either \emph{must} be used.
